Open libraries

```{r libraries, message = F}
library(rmongodb)
library(dplyr)
```

### Downloading datasets
The train usage statistics are downloaded as excel files and exported from excel to csv, then read into R below. To ensure that this operation works make sure the working directory has the exported csv files in it.
This comes from 2 sources:

  * Overground trains using Office of Ratio Regulation Report http://data.london.gov.uk/dataset/train-station-usage/resource/b23a39a1-f387-43ba-950a-19d1cfd8aab6.
    * This data is primarily based on ticket sales in each station and Oyster Card usage after January 2010. 
  * Underground data comes from TFL statistics http://www.tfl.gov.uk/cdn/static/cms/documents/multi-year-station-entry-and-exit-figures.xls.
    * There is less information in the metadata about how these figures were produced it is assumed that it is based on oyster data as well but there are both discrepencies and weightings/estimates taking place. 
    
* The positional data for the trains is from http://www.doogal.co.uk/london_stations.php 
    * This is a less reliable source because it isn't official but seems to be accurate. 
    * This has already been uploaded to mongodb so must be downloaded again to clean. (as it is a small dataset)


```{r downloading Data, results='hide'}
# Read overground train data 
trainUsage <- read.csv("station-usage.csv", stringsAsFactors = F)
# Read underground train data
tubeUsage <- read.csv("multi-year-station-entry-and-exit-figures.csv", skip=6, stringsAsFactors = F, strip.white = T)
stations <- read.csv("London stations.csv", stringsAsFactors = F)

#------ Defining gloabl variables -----------
DB <- "DSproject"
CRIME <- paste(DB, "londonCrime", sep=".")
STATIONS <- paste(DB, "stations", sep=".")


# Create mongo connection
if(exists("mongo")) try(mongo.destroy(mongo))
mongo <<- mongo.create(host = "mongodb-dse662l3.cloudapp.net:27017", db = DB)
```

```{r, eval=FALSE, echo=FALSE}
# Download stations as a dataframe
ifelse(mongo.is.connected(mongo), 
       lstations <- mongo.find.all(mongo, STATIONS),
       {
         createMongo()
         ifelse(mongo.is.connected(mongo),
                lstations <- mongo.find.all(mongo, STATIONS),
                stop("Could not connect to mongoDB"))
       })
stations <- data.frame(
  Station = sapply(lstations, function(x) x$Station),
  Latitude = sapply(lstations, function(x) x$Latitude),
  Longitude = sapply(lstations, function(x) x$Longitude),
  OSX = sapply(lstations, function(x) x$OS.X),
  OSY = sapply(lstations, function(x) x$OS.Y),
  Zone = sapply(lstations, function(x) x$Zone),
  Postcode = sapply(lstations, function(x) x$Postcode)
)

stations <- select(stations, -OS.X, - OS.Y)
```

### Cleaning the stations dataset
* Setting the type
* Fixing some anomalous naming (to match merging dataset)

```{r Cleaning geoStations}
# uploaded data is stored in strings so change the type of variables to be appropriate
stations$Latitude <- as.numeric(stations$Latitude)
stations$Longitude <- as.numeric(stations$Longitude)
stations$OSX <- as.numeric(stations$OS.X)
stations$OSY <- as.numeric(stations$OS.X)
stations$Zone <- as.factor(stations$Zone)
stations$Station <- as.character(stations$Station)

# Fix some discrepencies in station names
stations$Station <- gsub("^Acton$", "Acton Main Line", stations$Station)
stations$Station <- gsub("^Bank$", "Bank & Monument", stations$Station)
stations$Station <- gsub(" and ", " & ", stations$Station, perl = T)
stations$Station <- gsub("^Earls Court$", "Earl's Court", stations$Station)
stations$Station <- gsub("Bakerloo", "Bak", stations$Station)
stations$Station <- gsub("Circle/District/Hammersmith & City", "Cir", stations$Station)
stations$Station <- gsub("District", "Dis", stations$Station)
stations$Station <- gsub("Met.", "H&C", stations$Station)
stations$Station <- gsub("1 2 3", "123", stations$Station)
stations$Station <- gsub("^St Pancras$", "St.Pancras", stations$Station)
str(stations, strict.width = "cut")
```

### Cleaning Data
To clean the overground trains data select the relevant variables and the subset of the data in london. 
We then change the names of the variables to be more readable and turn the comma seperated usage numbers into numeric type.
```{r Cleaning overground trains}
# Select relevent observations and variables
trainUsage <- filter(trainUsage, Government.Office.Region..GOR. == "London")
trainUsage <- select(trainUsage, Station.Name, District.or.Unitary.Authority, Station.Facility.Owner, X1415.Entries...Exits, Explanation.of.large.change)
# Rename variables
names(trainUsage) <- c("Station", "Borough", "Owner", "TicketUsage", "railComments")
# Turn usage into a number (without commas)
trainUsage$TicketUsage <- as.numeric(gsub(",","", trainUsage$TicketUsage))
str(trainUsage, strict.width = "cut")
```
To clean the underground data we perform similar steps of selecting the relavent variables and filtering out some additional rows which are either empty or related to the "Total".
Then the usage numbers are multiplied by 1000000 because they are in millions. 
Finally the comments section is keyed so we replace the key values with the actual comments they refer to. 
```{r Cleaning underground trains}
# Select relevant variables
tubeUsage <- select(tubeUsage, Station, million, Note)
names(tubeUsage) <- c("Station", "EntryExitEst", "tubeComments")
# Remove some NA rows and "Total" row
tubeUsage <- filter(tubeUsage, !(Station == "Total"), !(Station == ""))
# Convert usage to units (instead of millions)
tubeUsage$EntryExitEst <- tubeUsage$EntryExitEst * 1000000
# Change the key of tubeComments into their actual value
tubeUsage$tubeComments <- gsub("^A$", "", tubeUsage$tubeComments)
tubeUsage$tubeComments <- gsub("^B$", "Figures for Bank/Monument now exclude W&C line passengers interchanging with other lines", tubeUsage$tubeComments)
tubeUsage$tubeComments <- gsub("^T$", "Figures include all passengers on the W&C line (which is closed on Sundays", tubeUsage$tubeComments)
tubeUsage$tubeComments <- gsub("^N$", "Station Closed", tubeUsage$tubeComments)
str(tubeUsage, strict.width = "cut")
```


### Merging trainUsage & Station 
We first simply merge the three datasets together over the station name variable. Note that now some of the fields are sparse because not all stations are in both dataset. 

We can observe that some stations have usage variables for both underground and rail datasets. 
  * this is primarily when the station has both an overground and underground component for example waterloo station. For these stations we have chosen to use the rail usage simply because more information was availible about the process involved and it is more likely that it has a more complete picture because it does take into account both paper  tickets and oyster card so definitely factors in the tube part of those stations, whereas it is unclear if this is the case the other way around.
We can observe some stations have no usage variables from either datasets.
  * This is for one of two reasons, naming discrepency between the location dataset and those trying to be merged; we attempted to account for this when cleaning these datasets above but some may still remain. Secondly the small stations for the DLR and the tramlink aren't included. We keep this data because it can still have a relevant effect on crime but the NA values must be accounted for by the visualisation process. 
Finally not all of the stations from the usage datasets are matched to those in the location datasets. These were stations that were quite far out of london so the curator of the location dataset considered it not to be a london station. As we do not have the location of these stations and it is questionable if they are in london or not we have chosen to ignore them (as is the default procedure of the script)
```{r merging}
mergedStations <- merge(stations, trainUsage, "Station", all.x = TRUE)
mergedStations <- (merge(mergedStations, tubeUsage, "Station", all.x = TRUE))
mergedStations <- mutate(mergedStations, usage = ifelse(!is.na(TicketUsage), TicketUsage, EntryExitEst))
str(mergedStations, strict.width = "cut")
```

```{r adding type}
#mergedStations$Owner <- 
  invisible(sapply(1:nrow(mergedStations), function(x) {
  ifelse(!is.na(mergedStations$EntryExitEst[x]) & is.na(mergedStations$TicketUsage[x]), {
    mergedStations$Owner[x] <<- "TFL"
    mergedStations$Type[x] <<- "Tube"
  },
  ifelse(is.na(mergedStations$EntryExitEst[x]) & !is.na(mergedStations$TicketUsage[x]),
  mergedStations$Type[x] <<- "Rail",
  ifelse(!is.na(mergedStations$EntryExitEst[x]) & !is.na(mergedStations$TicketUsage[x]),
  mergedStations$Type[x] <<- "Tube, Rail",
  mergedStations$Type[x] <<- "Other"
  )))
}))

```
Manually add the borough information by querying the website http://local.direct.gov.uk/LDGRedirect/index.jsp?mode=1.1
```{r adding borough, echo = FALSE}
index <- which(is.na(mergedStations$Borough))
boroughs <- c("Newham", "Ealing", "Croydon", "Croydon", "City of London", 
              "Tower Hamlets", "Tower Hamlets", "Brent", "Buckinghamshire", "Croydon", 
              "Islington", "Islington", "Croydon", "Enfield", "Islington", 
              "Bromley", "Westminster", "City of London", "Sutton", "City of London", 
              "Redbridge", "Hammersmith and Fulham", "Westminster", "Bromley", "Newham", 
              "Newham", "Barking and Dagenham", "Sutton", "Sutton", "Camden", 
              "Southwark", "Tower Hamlets", "Croydon", "Tower Hamlets", "Westminster", 
              "Southwark", "Hounslow", "Haringey", "Tower Hamlets",  "Tower Hamlets", 
              "Barnet", "Borough", "Tower Hamlets", "Broxbourne", "Epping Forest", 
              "Barnet", "Hertfordshire", "Islington", "Camden", "Tower Hamlets", 
              "Newham", "Harrow", "Hertforshire", "Surrey", "Croydon", 
              "Thurrock", "Buckinghamshire", "Camden", "Camden", "Buckinghamshire", 
              "Broxbourne", "Epping Forest", "Reigate and Banstead", "Ealing", "Heretfordshire", 
              "Croydon", "Lambeth", "Lambeth", "Wandsworth", "Elmbridge", 
              "Enfield", "Barnet", "Merton", "Croydon", "Westminster", 
              "Tower Hamlets", "Heretfordshire", "Newham", "Greenwich", "Newham", 
              "Barking and Dagenham", "Barking and Dagenham", "Hackney", "Dartford", "Epping Forest", 
              "Greenwich", "Tower Hamlets", "Brent", "Merton", "Ealing", 
              "Kensington and Chelsea", "Hammersmith and Fulham", "Barnet", "Newham", "Tower Hamlets", 
              "Wandsworth", "Hillingdon", "Barnet", "Westminster", "Westminster", 
              "Havering", "Hertfordshire", "Lewisham", "Westminster", "Epping Forest", 
              "Reigate and Banstead", "Camden", "Epsom and Ewell", "Epsom and Ewell", "Redbridge", 
              "Croydon", "Barnet", "Camden", "Hammersmith and Fulham", "Newham", 
              "Redbridge", "Croydon", "Kensington and Chelsea", "Barnet", "Hammersmith and Fulham", 
              "Camden", "Epping Forest", "Croydon", "Westminster", "Westminster",
              "Redbridge", "Hammersmith and Fulham", "Hammersmith and Fulham", "Camden", "Elmbridge",
              "Ealing", "Croydon", "Harrow", "Spelthorne", "Bromley", 
              "Hillingdon", "Hillingdon", "Hillingdon", "Barnet", "Tower Hamlets", 
              "Barnet", "Kensington and Chelsea", "Haringey", "Hillingdon", "Camden", 
              "Kensington and Chelsea", "Islington", "Havering", "Hounslow", "Hounslow",
              "Hounslow", "Westminster", "Hillingdon", "Tower Hamlets", "Southwark",
              "Hammersmith and Fulham", "Brent", "Brent", "Newham", "Croydon", 
              "Camden", "Brent", "Reigate and Banstead", "Kensington and Chelsea", "Kensington and Chelsea",
              "Lambeth", "Westminster", "Tower Hamlets", "Kensington and Chelsea", "Croydon",
              "Westminster", "Waltham Forest", "Waltham Forest", "Croydon", "Newham",
              "Epping Forest", "Westminster", "Hackney", "City of London", "Westminster",
              "Merton", "Tower Hamlets", "Barnet", "Merton", "City of London",
              "Hertfordshire", "Merton", "Merton", "Camden", "Tower Hamlets",
              "Brent", "Croydon", "Redbridge", "Ealing", "Ealing", 
              "Greenwich", "Harrow", "Ealing", "Ealing", "Brent",
              "Hillingdon", "Hillingdon", "Kensington and Chelsea", "Enfield", "Thurrock",
              "Hounslow", "Lambeth", "Westminster", "Ealing", "Hammersmith &  Fulham", 
              "Ealing", "Merton", "Westminster", "Westminster", "Harrow", 
              "Newham", "Newham", "Tower Hamlets", "Brent", "Newham",
              "Newham", "Thurrock", "Hammersmith and Fulham", "Brent", "Southwark",
              "Brent", "Lambeth", "Westminster", "Hertfordshire", "Havering",
              "Hammersmith and Fulham", "Harrow", "Redbridge", "Croydon", "Croydon",
              "Westminster", "Hertfordshire", "Epping Forest", "Newham", "Westminster",
              "Newham", "Hillingdon", "Hillingdon", "Hillingdon", "Camden",
              "Croydon", "Brentwood", "Hammersmith and Fulham", "Kensington and Chelsea", "Croydon",
              "Redbridge", "Ealing", "Harrow", "Kensington and Chelsea", "Tower Hamlets", 
              "Merton", "Redbridge", "Wandsworth", "Enfield", "Southwark", 
              "Merton", "Waltham Forest", "Lewisham", "Richmond upon Thames", "Bromley",
              "Westminster", "Westminster", "City of London", "Hounslow", "Harrow",
              "Newham", "Tower Hamlets", "Lambeth", "Epsom and Ewell", "Newham",
              "Harrow", "Brent", "Sutton", "Camden", "Reigate and Banstead",
              "Reigate and Banstead", "Westminster", "Elmbridge", "Broxbourne", "Croydon",
              "Epping Forest", "Wandsworth", "Wandsworth", "Westminster", "Barnet",
              "City of London", "Tower Hamlets", "Camden", "Hounslow", "Haringey",
              "Havering", "Barking and Dagenham", "Surrey", "Newham", "Hillingdon",
              "Croydon", "Broxbourne", "Croydon", "Redbridge", "Camden", 
              "Westminster", "Hertfordshire", "Hertfordshire", "Hertfordshire", "Croydon",
              "Brent", "Ealing", "Barnet", "Harrow", "Tower Hamlets", 
              "Hammersmith and Fulham", "Newham", "Kensington and Chelsea", "Tower Hamlets", "Westminster",
              "Hammersmith and Fulham", "Surrey", "Surrey", "Brent", "Merton", 
              "Haringey", "Hammersmith and Fulham", "Redbridge", "Croydon", "Barnet"
)

  mergedStations$Borough[index] <- boroughs

# Found an incorrect post code
mergedStations$Postcode <- as.character(mergedStations$Postcode)
mergedStations$Postcode[which(mergedStations$Station == "Bond Street")] <- "W1C 2JS"
mergedStations$Postcode[which(mergedStations$Station == "Canning Tower")] <- "E16 1ED"
mergedStations$Postcode[which(mergedStations$Station == "Deptford Bridge")] <- "SE8 4HH"
mergedStations$Postcode[which(mergedStations$Station == "Epsom Downs")] <- "KT17 4JP"
mergedStations$Postcode[which(mergedStations$Station == "Euston Square")] <- "NW1 2BF"
mergedStations$Postcode[which(mergedStations$Station == "George Street")] <- "CR0 1LA"
mergedStations$Postcode[which(mergedStations$Station == "Heathrow Terminal 4")] <- "TW6 3YD"
mergedStations$Postcode[which(mergedStations$Station == "Heron Quays")] <- "E14 4AD"
mergedStations$Postcode[which(mergedStations$Station == "Manor House")] <- "N4 2LX"
mergedStations$Postcode[which(mergedStations$Station == "Purfleet")] <- "RM19 1PL"
mergedStations$Postcode[which(mergedStations$Station == "Regents Park")] <- "NW1 5HD"
mergedStations$Postcode[which(mergedStations$Station == "Tottenham Court Road")] <- "W1D 1AP"
mergedStations$Postcode[which(mergedStations$Station == "Tufnell Park")] <- "N19 5AA"
mergedStations$Postcode[which(mergedStations$Station == "Wood Lane")] <- "W12 7TU"

```
Finally we replace our stations dataset on mongodb with the merged stations data. 
```{r upload, eval=FALSE}
mergedStations <- select(mergedStations, -OS.X, -OS.Y)
if(mongo.is.connected(mongo) == TRUE) {
  bson <- mongo.bson.from.df(mergedStations)
  mongo.drop(mongo, STATIONS)
  mongo.insert.batch(mongo, paste0(DB, ".stationsCrime"), bson) 
}
```