---
title: "StatisticsCoursework"
author: "Alun Meredith"
date: "19 October 2015"
output: pdf_document
---
```{r, dependencies}
library(ggplot2)
library(reshape2)
```

Downloaded fish.txt, data about the catch of a hypothetical fishing fleet from "http://www.edshare.soton.ac.uk/view/courses/COMP6235/2015.html" on `r date()`. 

```{r, read lines}
readLines("fish.txt", 5)
```

From reading the first 5 lines of the file we can see that the data is 2 columns of numerics. Each variable is seperated by a single whitespace, there is no header and no apparent NA characters or comment/escape characters. As such we can use the default values of read.table, including the col.names argument to name the variables.

```{r, import data}
fish <- read.table("fish.txt", col.names = c("times", "size"))
attach(fish)
```

Initial viewing of the data
```{r, step 1}
clearTheme <- theme(panel.background = element_blank(), 
  panel.grid.minor = element_blank(), 
  panel.grid.minor = element_blank()) 

top <- ggplot() +
  geom_density(aes(times), adjust=0.5) +
  clearTheme +
  geom_vline(xintercept = mean(times)) +
  geom_vline(xintercept = median(times)) + 
  geom_vline(xintercept = exp(mean(log(times))))

right <- ggplot() +
  geom_density(aes(size), adjust=0.5) + 
  coord_flip() +
  clearTheme +
  geom_vline(xintercept = mean(size)) +
  geom_vline(xintercept = median(size)) + 
  geom_vline(xintercept = exp(mean(log(size))))


scatter <- ggplot(data=fish) + 
  geom_point(aes(times,size)) 

empty <- ggplot() + 
  geom_point(aes(1,1), colour="white") +
  clearTheme +
  scale_x_discrete(labels="") +
  scale_y_discrete(labels="")

grid.arrange(top, empty, scatter, right, ncol=2, nrow=2, widths=c(4, 1), heights=c(1, 4))

```

```{r}
summary(times)
var(times)
skewness(times)
kurtosis(times)
exp(mean(log(times)))
```